Epidemiological dynamics in idealized metapopulations defined by asymmetric homophily

### 2024-07-14 index.qmd
updated index and restarted Rproj

---
title: Epidemiological dynamics in idealized metapopulations defined by asymmetric homophily
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    cite-method: biblatex
    biblatexoptions: sorting=none, style=numeric-comp
    pdf-engine: latexmk
    pdf-engine-opt: -xelatex
authors:
  - name: Matthew Adam Turner
    affiliation: Stanford University Doerr School of Sustainability
    roles: writing, model development
    corresponding: true
    email: maturner@stanford.edu
  - name: Izabela Rezende
  - name: Benjamin Singer
  - name: Hannah S. Derwanz
  - name: James Holland Jones
    affiliation: Stanford University Doerr School of Sustainability
    roles: writing
    corresponding: true
bibliography: ../library.bib
classoption: abstract
keep-tex: true
abstract: |
  This is the abstract
geometry:
  - margin=1.1in
  - letterpaper
biblio-title: References
---

@@

Due to rapidly increasing population size and density in a changing climate, the interdisciplinary science of pandemic preparedness is an increasing priority [@Rodo2021; @Adashi2022]. To complement our biological understanding and preventative tools for mitigating the risk of pandemics, social science is essential to understand how social structure and behavior determines how disease and disease prevention behaviors (such as vaccination[@Amin2017; @Omer2021]) are adopted by individuals in different group-structured populations [@Buckee2021; @Epstein2008; @SmaldinoJones2021]. Oftentimes in epidemiology, groups are defined by any different group partitions can be relevant to epidemiology, including age (RSV and Polio refs), occupation (frontline/remote refs), and, of course, political party [@VanBavel2020]. To understand how social structure affects . Using this we can infer the effect of different vaccine distribution allotments between the two groups that could possibly better protect one group or the other, or meet other optimal criteria. Computational analyses of metapopulation models can be used to create *in silico* experiments that would be impossible to implement in the real world, for instance to predict how different vaccination distribution strategies would best prevent transmission dynamics [@Singer2022].

Whether a new viral mutation or a cultural adaptation, beneficial *innovations* diffuse differently through different metapopulation structures. Turner and colleagues (2023) showed that novel innovations with higher fitness than an alternative best diffused when first introduced to a model minority group compared to a majority group (in terms of population size), in a metapopulation with a single other majority group [@Turner2023], and diffusion was further optimized when majority groups moderately preferred in-group interactions. Preference for in- versus out-group interactions is known as *homophily* [@McPherson2001], and introduced to a simple model metapopulation composed of a minority and majority group (by population size), where there exists a lower-fitness alternative, fitness-biased social learning.

Simple idealized models like we develop here serve as a theoretical ground truth against which we can better understand real-world complexities. For instance, schistosomaiasis hotspots can be predicted with some accuracy based on data from initial outbreak [@Singer2024], but metapopulation structure may have an underlying effect on the emergence of hotspots that, when accounted for, could increase predictive power of data-driven statistical models. Metapopulation structure with asymmetric homophily is not yet rigorously understood in an idealized theoretical setting, despite the fact that contact structure can be observed to affect transmission dynamics of sexually transmitted disease [@Morris1991], the spread of antibiotic-resistant bacteria across hospital wards [@Shapiro2020], the emergence and diffusion of drug-resistant malaria [@Ariey2003], the spread of schistosomiasis [@Ciddio2017], and provide disease reservoirs that drove the spread of tuberculosis and COVID-19 between some prison and community populations [@Sacchi2015; @Ryckman2021]. Contact structure has been rigorously shown to shape epidemiological dynamics [@Arthur2017; @Großmann2021], but not in settings where social networks are defined by asymmetric homophily, specifically. For data-driven applications, asymmetric homophily measurements may be easier or more reliable compared to the network structure itself.

To understand the effect of asymmetric homophily levels on epidemiological dynamics, we developed a computational model where metapopulations are composed of two groups, a minority and majority group defined by population size, where group members interact more or less frequently depending on group membership and group-level homophily levels, and whether interactions are structured by a social network (more like in a sexually-transmitted disease) or by random asymmetric homophilous mixing within and between groups (like in an airborne disease), adapting the models of Turner et al (2023) and Turner, Reynolds, and Jones (2024) for epidemiology by augmenting the social learning dynamic with SIR epidemiological dynamics.

Models like this enable us to rigorously understand how one important mechanism affects the course of disease and optimal intervention strategy. The cost of rigor, at this stage, is the exclusion of other important details that can affect disease or cultural transmission, including individual-level general health, access to clean running water, working conditions, and social factors such as asymmetric outgroup aversion levels between groups, which can undermine the adoption of adaptive behaviors [@EhretConstantino2022]. However, this initial mechanistic model ideally sets in motion a series of follow-on studies to better understand which of these additional factors alter model predictions to what extent. Model refinements and specializations can be added for specific epidemiological contexts, different scenarios and intervention strategies can be tested computationally for optimal policies given available information. More shortcomings, refinements, and specializations can follow that, with the process repeating as necessary. This work complements a broader effort across evolutionary sciences to understand the interplay between social structure and collective co-adaptation among humans, non-humans, and pathogens [@Bak-Coleman2021; @GalesicEtAl2023].

@@

@@

## References

>>>>

---
title: Epidemiological dynamics in idealized metapopulations defined by asymmetric homophily
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    cite-method: biblatex
    biblatexoptions: sorting=none, style=numeric-comp
    pdf-engine: latexmk
    pdf-engine-opt: -xelatex
authors:
  - name: Matthew Adam Turner
    affiliation: Stanford University Doerr School of Sustainability
    roles: writing, model development
    corresponding: true
    email: maturner@stanford.edu
bibliography: [../library.bib]
# bibliography: this.bib
classoption: abstract
keep-tex: true
abstract: |
  \noindent hello
geometry:
  - margin=1.1in
  - letterpaper
biblio-title: References
---

@@

To equitably prepare for pandemics in our uncertain world we need to [@Rodo2021; @Adashi2022]. The global COVID-19 pandemic highlighted the current lack of outcome equality, where a country's inequality (measured by the Gini index) predicted higher burden of COVID-19 infections [@Su2022]. Developing computational models can help us understand basic behavioral and social factors underlying the relationship between behavior and disease, but need to be applied to real-world situations because, first and foremost, there are vulnerable populations who need effective mitigation plans for the inevitable next pandemic. Second, a generic, yet predictive and application-ready, framework for behavioral-epidemiological metapopulation models can only be developed through making contact with real-world scenarios to hone the model, identifying missing or unnecessary components, and building a repertoire or menu of potential modeling choices that may be important. By applying abstract mechanistic models to high-priority real-world problems, we will develop a more broadly useful science of pandemic preparedness.

In this work we contribute to this program of more equitable pandemic preparedness by developing a compartmental behavioral-epidemiological metapopulation model of COVID-19 transmission between a Brazilian city, Dourados, in Mato Grosso do Sul, and the neighboring Dourados Indigenous Reserve [@DeOliveira2023]. De Oliveira and co-authors hypothesized that COVID-19 variant diffusion from Dourados City into the Dourados Reserve is driven by young Indigenous men traveling between the reserve and the city of Dourados for economic reasons. COVID-19 burden was felt disproportionately by Indigenous peoples in Brazil [@Simionatto2020], but may have been worse in the Dourados Reserve due to its immediate proximity to the Dourados. There are many complicating factors that make it difficult to evaluate this hypothesis directly without adding some additional qualifications. Here we develop a mechanistic model of epidemiological transmission with asymmetric social mixing that evaluates how best to model group structure in the Reserve. In turn, this model is used to evaluate potential economic interventions that target aid to different groups to reduce infections in the reserve.

-   De Oliveira and colleagues tracked the spread of different SARS-CoV-2 variants by village and by variant.

![Time series of COVID-19 spread through the Dourados Indigenous Reserve, with significant events highlighted. The first B.1.1 variant cases were introduced by the Guarani ethnicity (1), followed about one month later by a rapid rise in cases in Jaguapiru among the Terena ethnicity(2); several months later a rapid rise in Bororó was caused by a rapid rise among Guarani (3). Zeta was introduced at the end of 2020 (4), with a rapid rise around the holiday season especially in Jaguapiru among Guarani (5); later in February cases rose in Bororó among the Terena (6).](Figures/Time%20Series.pdf){width=80%}

-   The goal of our model is two-fold. First, to identify parameters that could re-create the time series plots in Figure 1.
-   Second, we aim to use a parameterized model to understand how different intervention strategies could reduce infections among Reserve residents. For example, perhaps since Terena are suspected to travel to the 
-   While we know a lot about transmission in the reserve, and the work of de Oliveira and colleagues advanced our understanding of transmission in the Reserve, we do not have reliable estimates of exactly how many cases there were per ethnicity and village. We also have only word-of-mouth information about the frequency of travel to the city for each ethnicity and location. Our model motivates the need for a more thorough understanding of behavior and epidemiological parameters in the Dourados Indigenous Reserve, and points the way towards future research in other Indigenous communities, and more complex situations.

@@

![We can model the relevant groups for mixing as either village residence or ethnicity. Transmission, mixing, and mobility parameters would be set accordingly. The $C_g$ parameters would be reduced if economic aid were allocated to keep group $g$ on the Reserve during outbreaks.](Figures/ModelSketch.pdf){fig-align="center"}

@@
